---
title: "Descriptive data analysis"
author: "Emine"
date: "2023-06-18"
output: html_document
---

```{r setup, include = TRUE, echo = FALSE}
knitr::opts_chunk$set(message = FALSE)
```

```{r}
here::i_am("Code/DescriptiveAnalysis.Rmd")
```


```{r, warning = FALSE, echo = TRUE}
library(tidyverse)
library(openxlsx)
library(data.table)
library(ggplot2)
library(geomtextpath)
library(psych)
library(cluster)
library(here)
```

# Descriptive Analysis 
Here we are going to use "subset_dt.xlsx" (I created excel instead of csv because
it is easier to work on tableau). This file created based on DataCleaning_exp.rmd. 

Let's first read the table and briefly check if everything is okay.  
And do not forget that this is sampling paradigm, which means there is no feedback!!!!!!


```{r, warning = FALSE, echo = TRUE}
#read the file 
main_dt <- as.data.table(read.xlsx(here("Data", "subset_dt.xlsx")))
colnames(main_dt)
```
```{r, warning = FALSE, echo = TRUE}
# Start with simple descriptive 

# Number of different session (identifier) 
total_identifier <- main_dt[, .N]

# Total number of participants 
total_participant <- main_dt[, uniqueN(participant)]

# Total number of trials 
total_trial <- main_dt[, sum(sampling_total)]

# Average number of trials by identifier 
avg_trial_identifier <- main_dt[, mean(sampling_total)]


# Average identifier by participant 
# which shows the average number of problems that each 
# participant solve 
avg_questions <- total_identifier/total_participant

```


# Did participant see all outcomes within the trial?

```{r}

total_seen_all <- main_dt[seen_all == "TRUE", .N]
percentage_of_seen_all <- round(total_seen_all/total_identifier, 3)

print(paste0("Only in ", percentage_of_seen_all*100, "% of the run, participants saw all the outcomes"))


```

# Did the participant see similar distribution 
Here I decided to use heuristic 0,1 difference in probabilities 

```{r}
main_dt[, seen_similar_distA := ifelse(abs(seen_probA1 - probA1) <= 0.1, TRUE, FALSE)]
main_dt[, seen_similar_distB := ifelse(abs(seen_probB1 - probB1) <= 0.1, TRUE, FALSE)]
main_dt[, seen_similar_dist := ifelse(seen_similar_distA == TRUE & seen_similar_distB == TRUE, TRUE, FALSE)]

total_similar_dist <- main_dt[seen_similar_dist == TRUE, .N]
percentage_of_seen_similar_dist <- round(total_similar_dist / total_identifier, 3) 

print(paste0(percentage_of_seen_similar_dist*100, "% of the run, participants saw similar distribution to the description"))

```

# How much of the participant see similar distribution and all outcmes? 

```{r}
main_dt[, seen_similar := ifelse(seen_similar_dist == TRUE & seen_all == TRUE, TRUE, FALSE)]
total_seen_similar<- main_dt[seen_similar == TRUE, .N]
percentage_of_seen_similar <- round(total_seen_similar / total_identifier, 3) 

print(paste0(percentage_of_seen_similar*100, "% of the run, participants saw similar distribution with all outcomes to the description"))
```

# sampling total basics

``````{r, warning = FALSE, echo = TRUE}

mean(main_dt$sampling_total)
median(main_dt$sampling_total)
max(main_dt$sampling_total)
min(main_dt$sampling_total)

```
Histogram for trials (sampling total)

```{r, warning = FALSE, echo = TRUE}

# Create a histogram with a bin size of 2 
# added median line to the graph

ggplot(main_dt, aes(x = sampling_total)) + geom_histogram(binwidth = 2, fill="black", col="grey") + 
  geom_vline(xintercept = median(main_dt$sampling_total), linetype = "dashed", color = "red", size = 1) + 
  labs(title = "Histogram of Sampling Total",
       x = "Sampling Total",
       y = "Frequency")
```

# Switching basics

```{r,  warning = FALSE, echo = TRUE}

mean(main_dt$sampling_total_switch)
median(main_dt$sampling_total_switch)
max(main_dt$sampling_total_switch)
min(main_dt$sampling_total_switch)
```

Histogram for switchings (sampling total)

```{r, warning = FALSE, echo = TRUE}

# Create a histogram with a bin size of 2 
# added median line to the graph

ggplot(main_dt, aes(x = sampling_total_switch)) + geom_histogram(binwidth = 2, fill="black", col="grey") + 
  geom_vline(xintercept = median(main_dt$sampling_total_switch), linetype = "dashed", color = "red", size = 1) + 
  labs(title = "Histogram of Sampling Total",
       x = "Switching Total",
       y = "Frequency")
```

# Switching ratio 

# Create a histogram with a bin size of 
# added median line to the graph

```{r}
# Switch ratio definition 
# cleaned_exp[, sampling_switch_ratio := round(sampling_total_switch/(sampling_total-1), 3), by = identifier]

# Based on the definition, if the sampling total is equal to 1, the switch ratio will be NA 

# Create a histogram 
# added median line to the graph

# notice that I didn't include NA swtch ratio to median calculation

ggplot(main_dt, aes(x = sampling_switch_ratio)) + geom_histogram( fill="black", col="grey") + 
  geom_vline(xintercept = median(main_dt$sampling_switch_ratio, na.rm = TRUE), linetype = "dashed", color = "red", size = 1) + 
  labs(title = "Histogram of switching ratio",
       x = "Switching ratio",
       y = "Frequency")


```

# Relationship between sampling total (trials) and total switch 
``````{r, warning = FALSE, echo = TRUE}

ggplot(main_dt, aes(x = sampling_total, y = sampling_total_switch)) + geom_point() + 
  geom_textvline(label = "median of sampling total", xintercept = median(main_dt$sampling_total), color = "red") + 
  geom_texthline(label = "median of total switch", yintercept = median(main_dt$sampling_total_switch), color = "red") +
  geom_textabline(label = "median of switch ratio", slope = median(main_dt$sampling_switch_ratio, na.rm = TRUE), color = "red")
```
# Let's also check round ratio 

```{r}
# Round ratio definition 
# cleaned_exp[, sampling_switch_ratio := round(sampling_round_total/(sampling_total-1), 3), by = identifier]

main_dt[,sampling_round_ratio := round(sampling_round_total/(sampling_total-1),3)]

# Based on the definition, if the sampling total is equal to 1, the switch ratio will be NA 

# Create a histogram 
# added median line to the graph

# notice that I didn't include NA swtch ratio to median calculation

ggplot(main_dt, aes(x = sampling_round_ratio)) + geom_histogram( fill="black", col="grey") + 
  geom_vline(xintercept = median(main_dt$sampling_round_ratio, na.rm = TRUE), linetype = "dashed", color = "red", size = 1) + 
  labs(title = "Histogram of Round ratio",
       x = "Round ratio",
       y = "Frequency")


```


```{r}


ggplot(main_dt, aes(x = sampling_round_ratio)) + geom_histogram( fill="black", col="grey") + 
  geom_vline(xintercept = median(main_dt$sampling_round_ratio, na.rm = TRUE), linetype = "dashed", color = "red", size = 1) + 
  labs(title = "Histogram of Sampling Total",
       x = "Round ratio",
       y = "Frequency")

ggplot(main_dt, aes(x = sampling_total, y = sampling_round_total)) + geom_point() + 
  geom_textvline(label = "median of sampling total", xintercept = median(main_dt$sampling_total), color = "red") + 
  geom_texthline(label = "median of total round", yintercept = median(main_dt$sampling_round_total), color = "red") +
  geom_textabline(label = "median of round ratio", slope = median(main_dt$sampling_round_ratio, na.rm = TRUE), color = "red") +
  geom_textabline(label = "mean of round ratio", slope = mean(main_dt$sampling_round_ratio[is.finite(main_dt$sampling_round_ratio)]), color = "red")
```

# SAMPLING STRATEGIES 
## Let's try to find out the groups of sampling strategies based on sampling related columns 

```{r}

# the columns that I will work on : sampling_total, sampling_switch_ratio, "sampling_round_ratio

# Select the columns of interest
sampling_dt <- main_dt[is.finite(sampling_round_ratio), c("sampling_total", "sampling_switch_ratio", "sampling_round_ratio")]
sampling_dt2 <- main_dt[is.finite(sampling_round_ratio), c("sampling_total", "sampling_switch_ratio", "sampling_round_ratio", "sampling_total_switch", "sampling_round_total")]

#standardized data since k-means use euclidian distance 
scale_sampling_dt <- scale((sampling_dt))

# Perform k-means clustering with 3 clusters
k <- 2
kmeans_result1 <- kmeans(scale_sampling_dt, centers = k)

# Add the cluster assignments as a new column in the data table
sampling_dt2$cluster <- as.factor(kmeans_result1$cluster)

# visualize the findings

ggplot(sampling_dt2, aes(x = sampling_total, y = sampling_total_switch, color = cluster)) + geom_point()

ggplot(sampling_dt2, aes(x = sampling_total, y = sampling_round_total, color = cluster)) + geom_point()


```
```{r}

# the columns that I will work on : sampling_total, sampling_total_switch",  "sampling_round_total

# Select the columns of interest
sampling_dt1 <- main_dt[is.finite(sampling_round_ratio), c("sampling_total", "sampling_total_switch", "sampling_round_total")]
sampling_dt2.1 <- main_dt[is.finite(sampling_round_ratio), c("sampling_total", "sampling_switch_ratio", "sampling_round_ratio", "sampling_total_switch", "sampling_round_total")]

#standardized data since k-means use euclidian distance 
scale_sampling_dt1 <- scale((sampling_dt1))

# Perform k-means clustering with 3 clusters
k <- 2
kmeans_result2 <- kmeans(scale_sampling_dt1, centers = k)

# Add the cluster assignments as a new column in the data table
sampling_dt2.1$cluster <- as.factor(kmeans_result2$cluster)

# visualize the findings

ggplot(sampling_dt2.1, aes(x = sampling_total, y = sampling_total_switch, color = cluster)) + geom_point()

ggplot(sampling_dt2.1, aes(x = sampling_total, y = sampling_round_total, color = cluster)) + geom_point()


```

```{r}

# the columns that I will work on : sampling_total, sampling_total_switch",  "sampling_round_total, sampling_round_ratio, sampling_switch_ratio

# Select the columns of interest
sampling_dt3.0 <- main_dt[is.finite(sampling_round_ratio), c("sampling_total", "sampling_switch_ratio", "sampling_round_ratio", "sampling_total_switch", "sampling_round_total")]
sampling_dt3.1 <- main_dt[is.finite(sampling_round_ratio), c("sampling_total", "sampling_switch_ratio", "sampling_round_ratio", "sampling_total_switch", "sampling_round_total")]

#standardized data since k-means use euclidian distance 
scale_sampling_dt3 <- scale((sampling_dt3.0))

# Perform k-means clustering with 3 clusters
k <- 2
kmeans_result3 <- kmeans(scale_sampling_dt3, centers = k)

# Add the cluster assignments as a new column in the data table
sampling_dt3.1$cluster <- as.factor(kmeans_result3$cluster)
main_dt[is.finite((sampling_round_ratio)), sampling_cluster := sampling_dt3.1$cluster]

# visualize the findings

ggplot(sampling_dt3.1, aes(x = sampling_total, y = sampling_total_switch, color = cluster)) + geom_point()

ggplot(sampling_dt3.1, aes(x = sampling_total, y = sampling_round_total, color = cluster)) + geom_point()


```

I decided to go with the last kmeans clusters. 

  1: low switchers
  
  2:high switchers


# DECISION STRATEGIES
Here we will use columns related to "seen" and try to identify decision strategies

  - Did they choose highest ev?
  - Did their choice strategy influenced by domain of the question?

```{r}
# let's add one more variable called seen_exval_chosen 
# it will return if the choice is aligned exval, false otherwise 

main_dt[, seen_exval_chosen := ifelse(is.null(seen_exval), NA, ifelse(choice == seen_exval, TRUE, FALSE)), by = identifier]

```

- Did they choose highest ev?
```{r}
table_dt1 <- main_dt[, .N, by = c("sampling_cluster", "seen_exval_chosen")][order(seen_exval_chosen)][order(sampling_cluster)]

table_dt1[, percent := N / sum(N) * 100, by = sampling_cluster]

table_dt1
```

- Did their choice strategy influenced by domain of the question?
```{r}

table_dt2 <- main_dt[, .N, by = c("dom", "sampling_cluster", "seen_exval_chosen")][order(seen_exval_chosen)][order(dom)][order(sampling_cluster)]

table_dt2[, percent := N / sum(N) * 100, by = c("sampling_cluster", "dom")]

table_dt2

```



# SAMPLING AND DECISION TOGETHER 
Here ww will try to identify the relationship between sampling and decision strategies 






